---
title: 'gamlss2.ISD: Modeling of size spectra under the GAMLSS framework'
author: "Alexis Lazaris"
output: github_document
bibliography: bibl.bib
nocite: '@*'
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Overview

The goal of gamlss2.ISD is to estimate the exponent of the Individual Size Distribution for size spectra as described with a bounded Power-Law distribution. The exponent can be estimated from raw body size data within the GAMLSS framework as applied by the `gamlss2` package. All the functionalities of the `gamlss2` package can be used. 

# Installation

The `gamlss2.ISD` package requires installation of `gamlss2`:
```{r, eval = FALSE}
install.packages("gamlss2",
  repos = c("https://gamlss-dev.R-universe.dev",
            "https://cloud.R-project.org"))
```

It can then be installed using `devtools`:
```{r, eval = FALSE}
devtools::install_github("alexislaz/gamlss2.ISD")
```


For the examples presented here, the `sizeSpectra` package is also needed:
```{r, eval = FALSE}
remotes::install_github("andrew-edwards/sizeSpectra")
```


# Usage
The main function of the `gamlss2.ISD` package is `bPL` that can be used inside a `gamlss2` call:

```{r use, eval = FALSE}
gamlss2(size ~ fac + s(x1, x2) + s(x3, by = fac) + s(id, bs = "re"),
        data = data,
        family = bPL(data$counts, data$size_minima, data$size_maxima))
```


# Examples

First load relevant packages:
```{r load, message = FALSE, warning = FALSE, include = TRUE}
library(dplyr)
library(ggplot2)
library(gamlss2)
library(sizeSpectra)

library(brms)
#devtools::install_github("jswesner/isdbayes")
library(isdbayes)

library(gamlss2.ISD)
```


## Estimate the ISD exponent from a single sample/population 

We can simulate a dataset of body sizes from the bounded Power-Law distribution using the `rPLB` from the `sizeSpectra` package:

```{r ex1_r, message = FALSE, warning = FALSE, include = TRUE}
d = data.frame(size = rPLB(1000, -2, 1, 1000))  # 1000 body sizes from ISD with b = -2 bounded in [1, 1000]
```

And then estimate the exponent using GAMLSS:

```{r ex1_m, message = FALSE, warning = FALSE, include = TRUE, results = "hide"}
m = gamlss2(size ~ 1, 
            data = d, 
            family = bPL(1, 1, 1000)) # all counts = 1, range [1, 1000] as simulated
```

```{r ex1_m_summary, message = FALSE, warning = FALSE, include = TRUE}
summary(m)
```


## Estimate the ISD exponent from multiple populations

As a more complicated example, we can simulate body sizes for different populations with different characteristics:

```{r ex2_r, message = FALSE, warning = FALSE, include = TRUE}
# population-specific size ranges
pop1_size_range = c(10, 800)
pop2_size_range = c(20, 1000)
pop3_size_range = c(1, 300)

# population-specific 'b'
pop1_b = -1.2
pop2_b = -1.5
pop3_b = -1.8

dat = rbind(
  data.frame(pop = "pop1", 
             size = round(rPLB(1e4, pop1_b, pop1_size_range[1], pop1_size_range[2])),
             mn = pop1_size_range[1], mx = pop1_size_range[2]),
  data.frame(pop = "pop2", 
             size = round(rPLB(1e4, pop2_b, pop2_size_range[1], pop2_size_range[2])),
             mn = pop2_size_range[1], mx = pop2_size_range[2]),
  data.frame(pop = "pop3", 
             size = round(rPLB(1e4, pop3_b, pop3_size_range[1], pop3_size_range[2])),
             mn = pop3_size_range[1], mx = pop3_size_range[2])
)
dat$pop = factor(dat$pop) # convert to factor for models
```

And then estimate the exponents (b) for each population:

```{r ex2_m1, message = FALSE, warning = FALSE, include = TRUE, results = "hide"}
m = gamlss2(size ~ pop, 
            data = dat, 
            family = bPL(1, dat$mn, dat$mx))
```

Timing:
```{r ex2_m1_time, message = FALSE, warning = FALSE, include = TRUE}
summary(m)$elapsed # seconds of model fitting
```

See the estimates:

```{r ex2_m1_summary, message = FALSE, warning = FALSE, include = TRUE}
predict(m, newdata = data.frame(pop = c("pop1", "pop2", "pop3")), se.fit = TRUE)
```


Alternatively, for faster fitting, we can aggregate equal sizes with their respective abundance:

```{r ex2_m2, message = FALSE, warning = FALSE, include = TRUE, results = "hide"}
dat_counts = dat |> 
  group_by(pop, size, mn, mx) |>
  summarize(n = n()) |>
  ungroup()

m = gamlss2(size ~ pop, 
            data = dat_counts, 
            family = bPL(dat_counts$n, dat_counts$mn, dat_counts$mx))
```

Reduced model fitting time:
```{r ex2_m2_time, message = FALSE, warning = FALSE, include = TRUE}
summary(m)$elapsed
```

Identical estimates:
```{r ex2_m2_summary, message = FALSE, warning = FALSE, include = TRUE}
predict(m, newdata = data.frame(pop = c("pop1", "pop2", "pop3")), se.fit = TRUE)
```

We can, also, utilize the random effects structure that `gamlss2` offers:

```{r ex2_m3, message = FALSE, warning = FALSE, include = TRUE, results = "hide"}
m = gamlss2(size ~ s(pop, bs = "re"), 
            data = dat_counts, 
            family = bPL(dat_counts$n, dat_counts$mn, dat_counts$mx))
```

Where we obtain similar estimates:
```{r ex2_m3_summary, message = FALSE, warning = FALSE, include = TRUE}
predict(m, newdata = data.frame(pop = factor(levels(m$model$pop), levels(m$model$pop))), se.fit = TRUE)
```

And we can, also, visualize using the built-in `plot.gamlss2`:

```{r ex2_m3_plot, message = FALSE, warning = FALSE, include = TRUE}
plot(m)
```


## Estimate the effect of a predictor on the ISD exponent

Finally, as a more complex example, we can simulate a predictor that affects the exponent of the ISD while using random intercepts:

```{r ex3_r, message = FALSE, warning = FALSE, include = TRUE}
# setup global intercept and slope and a hardcoded 'random' intercept
set.seed(1)

intercept = -1.5   # global
slope = 0.1        # global

x.lvls = seq(-1, 1, length.out = 20)  # 'x' covariate
g.lvls = factor(c("pop1", "pop2"))    # different samples

# setup grid
x = rep(x.lvls, each = length(g.lvls))
g = rep(g.lvls, length(x.lvls))

b0 = c(0.15, -0.2) # 'random' intercept deviations from global
# the random intercepts are:
intercept + b0

X = model.matrix( ~ x)
Z0 = model.matrix( ~ -1 + g)

y = (X %*% c(intercept, slope)) + (Z0 %*% b0) + rnorm(length(x), 0, 0.01)

# collect in a data.frame
dat = data.frame(y = y, x = x, g = g)
```

Visualize the dependence of the exponent on 'x':
```{r ex3_plot, message = FALSE, warning = FALSE, include = TRUE}
ggplot(dat) + 
  theme_classic(base_size = 15) + 
  geom_point(aes(x = x, y = y, colour = g)) + 
  geom_smooth(aes(x = x, y = y, colour = g), method = "lm", se = FALSE) +
  labs(x = "x", y = "ISD exponent", colour = "population")
```


And now simulate actual individual sizes from the grid of exponents:
```{r ex3_sim, message = FALSE, warning = FALSE, include = TRUE}
dat_sizes = do.call(rbind,
                    Map(function(beta, x, g) 
                      data.frame(g = g, 
                                 x = x, 
                                 size = rPLB(500, beta, 1, 1000)), 
                      dat$y, dat$x, dat$g))
```

And visualize our simulated sizes. We note that "pop2" has a steeper size spectra "slope" (b) and, hence, more smaller individuals:
```{r ex3_sim_plot, message = FALSE, warning = FALSE, include = TRUE}
ggplot(dat_sizes) + 
  theme_classic(base_size = 15) + 
  facet_wrap(vars(g)) + 
  geom_histogram(aes(x = size))
```


Fit the mixed model:
```{r ex3_m, message = FALSE, warning = FALSE, include = TRUE, results = "hide"}
# run model
m = gamlss2(size ~ x + re(fixed = ~x, random = ~1|g), data = dat_sizes, family = bPL(1, 1, 1000))
```

Inspect the model fit:

```{r ex3_m_diagnostics, message = FALSE, warning = FALSE, include = TRUE}
plot(m)  # built-in diagnostics
```


And retrieve the estimated intercepts and slopes:

```{r ex3_m_coefs, message = FALSE, warning = FALSE, include = TRUE}
# the estimated intercepts are:
coef(specials(m)$model)[, 1] + coef(m)[1]
# the actual intercepts are:
intercept + b0

# the estimated slopes are:
coef(specials(m)$model)[, 2] + coef(m)[2]
# the actual slope is:
slope
```


## Estimate the ISD exponent from real-world data

We'll use the built-in `IBTS_data` dataset of the `sizeSpectra` package with records of individual length measurements of fish species as described in @edwards2020. First, we can replicate the results from Figure 8 from that publication, using the GAMLSS framework. We also, examine a model that accounts for species-specific differences of the exponent of the Power-Law distribution. We, also, fit a model using `isdbayes`, which, equivalently fits a bounded Power-Law (truncated Pareto) under a bayesian framework (@wesner2024).


First, format the dataset:
```{r ibts_repl_fmt, message = FALSE, warning = FALSE, include = TRUE}
dat = IBTS_data |>
  group_by(Year) |>
  mutate(mn = min(bodyMass), mx = max(bodyMass)) |> # in Edwards et al. 2020, min and max sizes are separated by year analysis
  mutate(Year = factor(Year), SpecCode = factor(SpecCode)) |>
  ungroup()
```

Fit the GAMLSS models. We use the `binning = TRUE` argument of `gamlss2_control` to fit model more efficiently. First, a model only with a random effect for Year:
```{r ibts_repl_gamlss_year, message = FALSE, warning = FALSE, include = TRUE, results = "hide"}
m_gam_year = gamlss2(bodyMass ~ s(Year, bs = "re"), 
                data = dat, 
                family = bPL(dat$Number, dat$mn, dat$mx), 
                control = gamlss2_control(binning = TRUE)) 
```


And we, also, fit a model accounting for a simple inter-species effect on 'b':

```{r ibts_repl_gamlss_year_species, message = FALSE, warning = FALSE, include = TRUE, results = "hide"}
m_gam_year_species = gamlss2(bodyMass ~ s(Year, bs = "re") +  s(SpecCode, bs = "re"), 
                             data = dat, 
                             family = bPL(dat$Number, dat$mn, dat$mx), 
                             control = gamlss2_control(binning = TRUE))
```


And finally, collect predicted effects from both models (in a similar notation to `sizeSpectra`):

```{r ibts_repl_gamlss_eff, message = FALSE, warning = FALSE, include = TRUE}
eff_gam_year = m_gam_year$results$effects[[1]] |>
  mutate(b = fit + coef(m_gam_year),
         confMin = lower + coef(m_gam_year),
         confMax = upper + coef(m_gam_year)) 
eff_gam_year_species = m_gam_year_species$results$effects[[1]] |>
  mutate(b = fit + coef(m_gam_year_species),
         confMin = lower + coef(m_gam_year_species),
         confMax = upper + coef(m_gam_year_species))
```


Also, fit the `brm` model, using the truncated Pareto parameterized with Stan in the `isdbayes` package:
```{r ibts_repl_brm, message = FALSE, warning = FALSE, include = TRUE, results = "hide"}
m_bayes_year = brm(bodyMass | vreal(Number, mn, mx) ~ (1|Year), 
                   data = dat,
                   stanvars = stanvars,
                   family = paretocounts(),
                   chains = 1, iter = 1000)  # need to increase `chains` and `iter` for better estimates
```


And, once again, collect the effects of the `brm` fit (in a similar notation to `sizeSpectra`):
```{r ibts_repl_brm_eff, message = FALSE, warning = FALSE, include = TRUE}
eff_bayes_year = ranef(m_bayes_year)[[1]][, , "Intercept"] |>
  as.data.frame() |>
  mutate(b = Estimate + fixef(m_bayes_year)[, "Estimate"],
         confMin = Q2.5 + fixef(m_bayes_year)[, "Estimate"],
         confMax = Q97.5 + fixef(m_bayes_year)[, "Estimate"]) |>
  add_rownames("Year")
```


Finally, replicate the results of Figure 8 from @edwards2020. (NOTE: `sizeSpectra::eightMethods.count` creates figures by default in the home directory! take care to delete them afterwards!)
```{r ibts_repl_mle, message = FALSE, warning = FALSE, include = TRUE}
# replicate the MLE of 'b' from Edwards et al. 2020 using `eightMethods.count` 
eff_mle = do.call(rbind, 
                lapply(split(IBTS_data, IBTS_data$Year), 
                       function(data) subset(eightMethods.count(data, data$Year[1]), Method == "MLE")))
```

And plot all the estimates. After plotting we can see the almost identical estimates between GAMLSS, the bayesian fit, and the MLE from Edwards et al. 2020. Accounting for the different species, though, changes the estimated exponents.
```{r ibts_repl_plot, message = FALSE, warning = FALSE, include = TRUE, fig.dim = c(20, 12)}
# collect both estimations in a single dataset and plot predictions of 'b'
rbind(cbind(method = "gamlss - only year", eff_gam_year[, c("Year", "b", "confMin", "confMax")]), 
      cbind(method = "gamlss - year and species", eff_gam_year_species[, c("Year", "b", "confMin", "confMax")]),
      cbind(method = "brm - only year", eff_bayes_year[, c("Year", "b", "confMin", "confMax")]),
      cbind(method = "Edwards 2020", eff_mle[, c("Year", "b", "confMin", "confMax")])) |>
  ggplot() + 
  theme_classic(base_size = 15) + 
  geom_point(aes(x = Year, y = b, colour = method), position = position_dodge(width = 0.75)) + 
  geom_errorbar(aes(x = Year, ymin = confMin, ymax = confMax, colour = method), width = 0.0, position = position_dodge(width = 0.75))
```


Besides this simple random year intercept model, we can, also, fit a more complex model using the `IBTS_data` dataset. First clean-up the data a bit:
```{r ibts_new, message = FALSE, warning = FALSE, include = TRUE}
species_selection = c(105865, 126417, 126436, 127139, 126461, 126484, 127023, 127160, 105923) # select a few species
dat = IBTS_data |>
  group_by(SpecCode) |> 
  mutate(mn = min(bodyMass), mx = max(bodyMass)) |> # add minimum and maximum size by species (and not by Year)
  ungroup() |>
  filter(SpecCode %in% species_selection) |> 
  mutate(species = factor(SpecCode))
```

Now fit a more sophisticated GAM:
```{r ex4_m, message = FALSE, warning = FALSE, include = TRUE, results = "hide"}
m = gamlss2(bodyMass ~ species + s(Year, by = species, k = 5), 
            data = dat, 
            family = bPL(dat$Number, dat$mn, dat$mx),
            control = gamlss2_control(binning = TRUE))
```


And visualize the effects:

```{r ex4_eff, message = FALSE, warning = FALSE, include = TRUE}
prediction_grid = expand.grid(Year = sort(unique(m$model$Year)),
                              species = factor(levels(m$model$species), levels(m$model$species)))
predictions = cbind(prediction_grid, 
                    predict(m, 
                            newdata = prediction_grid, se.fit = TRUE,
                            terms = c("(Intercept)", "Year", "species"),
                            nogrep = FALSE))
predictions$species = specCodeNames$species[match(as.character(predictions$species), specCodeNames$speccode)] # use species scientific names

ggplot(predictions) + 
  theme_classic(base_size = 15) +
  facet_wrap(vars(species), scales = "free_y") + 
  geom_line(aes(x = Year, y = fit)) + 
  geom_ribbon(aes(x = Year, ymin = fit-se, ymax = fit+se), alpha = 0.2) + 
  labs(y = "b")

```




## References


